The Multi-Stencil Language

### 2016-01-21 ARTICLES/IJPP/article/formalism.tex
article IJPP Helene 21/01/2016

To numerically solve a set of PDEs, iterative methods (finite difference, finite volume or finite element methods) are frequently used to approximate the solution through a discretized (step by step) phenomena. Thus, the continuous time and space domains are discretized so that a set of numerical computations are iteratively (time discretization) applied onto a mesh (space discretization). In other words, the PDEs are transformed to a set of numerical computations applied at each time step on elements of the discretized space domain. Among those numerical computations is found a set of numerical schemes, also called \textit{stencil computations}, and a set of auxiliary computations also needed to perform the simulation, and also called \emph{local computations}.
This section gives formal definitions of what we call a \textit{stencil program} and its computations. Then, the different mid-grain parallelization techniques which can be applied on such program, are presented.

@@

\begin{mydef}
\textit{A mesh is a connected undirected graph $\mathcal{M}=(V,E)$, where $V\subset \Omega$ is the set of vertices and $E\subseteq V^2$ the set of edges. The set of edges $E$ of a mesh $\mathcal{M}=(V,E)$ does not contain bridges. It is said that the mesh is applied onto $\Omega$}.
\end{mydef}
\begin{figure}[!h]\begin{center}
  \resizebox{8cm}{!}{\includegraphics{./images/maillages.pdf}}
  \caption{From left to right, Cartesian, curvilinear and unstructured meshes.}
  \label{fig:mesh}
\end{center}\end{figure}
\begin{mydef}
\textit{The dimension of a mesh $\mathcal{M}=(V,E)$ applied onto $\Omega=\mathbb{R}^n$ is denoted $dim(\mathcal{M})=n$.}
\end{mydef}
A mesh can be structured (as Cartesian or curvilinear meshes), unstructured, regular or irregular (without the same topology for each element) and hybrid as illustrated in Figure~\ref{fig:mesh}.

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item \textit{An entity $e$ of a mesh $\mathcal{M}=(V,E)$ is defined as a subset of its vertices and edges, $e\subset (V\cup E)$.
\item A mesh entities group is denoted $E$ such that $E \in \mathcal{P}(V\cup E)$. It represents a set of entities of the same type.
\item Finally the set of mesh entities groups of a simulation is denoted $\mathcal{E}$.}
\end{itemize}
%\end{mydef}

@@

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item \textit{We denote a scalar as an identifier associated to a numerical value and applied onto a mesh $\mathcal{M}$ where $dim(\mathcal{M})=0$. In other words a scalar can be seen as a variable containing a numerical value. The set of scalars is denoted $\mathcal{S}$.}
\item \textit{$\mathcal{T}=\mathbb{R}$ is the continuous time domain of a numerical simulation.}
\item \textit{The discretization of the continuous time domain $\mathcal{T}$ is denoted as the pair $T(\Delta t,conv)$.} 
\begin{itemize}
\item \textit{$\Delta t \in \mathcal{S}$ represents the time interval of the numerical simulation, such that for a current time iteration $t_i$, the next time iteration is $t_{i+1} = t_i + \Delta t$. The default value of $\Delta t$ is $1$.
\item $conv:\mathcal{S}^n \rightarrow bool$ is a function which returns a boolean from a set of scalar variables. This function represents the convergence criteria of the simulation. 
\item At a given time step, the convergence criteria is evaluated such that if $conv$ returns $true$ the next time step can start such that $t_{i+1} = t_i + \Delta t$, otherwise the simulation ends. For example, the simplest $conv$ is typically to have two scalars: $t$ the time, and a fix number of iterations denoted $it$. At each time step $\Delta t=1$, and $conv$ returns the boolean expression $t<it$.}
\end{itemize}
\end{itemize}
%\end{mydef}

@@

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item \textit{A quantity is a function $\delta: E_{\delta} \mapsto V_{\delta}$ which associates each entity of a group $e\in E_{\delta}$ to a value $v\in V_{\delta}$ where $V_{\delta}$ is a data type, typically $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.
\item The set of quantities applied onto the mesh is denoted $\Delta$.
\item In the rest of this paper, the group of mesh entities on which a quantity $\delta$ is mapped is denoted $entity(\delta)=E_{\delta}$.}
\end{itemize}
%\end{mydef}

@@

In this section are considered two different types of computations, a reduction or a numerical kernel.

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item \textit{A computation domain $D$ is a subpart  a mesh entities group, $D \subset E \in \mathcal{E}$.
\item The set of computation domains of a numerical simulation is denoted $\mathcal{D}$.
\item A neighborhood $n$ is a function which for a given entity $e \in E_i$, returns a set of $m$ entities in $E_j$, $n : E_i \rightarrow E_j^m$. One can notice that $i = j$ is possible.
\item The set of neihborhood functions in a numerical simulation is denoted $\mathcal{N}$.
\item A single data is written during a computation and is denoted $w \in \Delta$. Thus, the computation domain of the computation is $D \subset E_w=entity(w)$.
\item A set of data to read and use in a computation is denoted $R$. Each element of $R$ is a tuple $(r,n)$, where $r \in \Delta$ and $n$ is a neighborhood function such that $n : E_w \rightarrow entity(r)^m$. The neighborhood indicates which entities, on which $r$ is applied, will be read during the computation.
\item A numerical expression $\text{exp}_{w}: entity(w) \times R^n \rightarrow V_{w}$ is a function which computes the values of the written quantity $w \in \Delta$ for a given entity $e \in entity(w)=E_w$, using a set of input data $R$.
\item A computation kernel $k$ of a numerical simulation is defined as $c(R,w,\text{exp},D)$, where $R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.}
\end{itemize}
%\end{mydef}
%If the number of computations in $\Gamma$ is $card(\Gamma)=n$, such that $\bigcup_{i=0}^{n-1}c_i = \Gamma$, then $\bigcup_{i=0}^{m-1}R_i \cup w_i \subseteq \Delta$.

@@

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item \textit{We denote by $identity$ the identity function $x=x$.
\item A kernel computation $k(R,w,\text{exp},D) \in \Gamma$ is a stencil kernel $\iff \exists (r,n) \in R$ such that $n \neq identity$.
\item A kernel computation $k(R,w,\text{exp},D) \in \Gamma$ is a local kernel $\iff \forall (r,n) \in R$, $n = identity$.}
\end{itemize}

\noindent \textbf{Property}
\textit{A kernel for which all data read and written are applied onto a mesh of dimension $0$ is a local kernel.}

@@

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item \textit{A reduction computation $r$ of a numerical simulation is defined as $r(R,w,red)$, where $R$ is the set of data read, $w \in \mathcal{S}$ is the scalar written, and $\text{red}$ is an associative reduction function.
% \item For any $i,j$ such that $r_i \in R$ and $r_j \in R$, if $dim(mesh(r_i))=dim(mesh(r_j))=n$.
% \item The data written by the reduction $w$ verifies $dim(mesh(w))<n$.
\item A reduction function $\text{red}: \Delta^n \rightarrow \mathcal{S}$ is an operation which reduces a set of data in $\Delta$ (on it sentities) to a single scalar variable in $\mathcal{S}$. The reduction function must be associative.
\item The set of $n$ ordered computations (kernel or reduction) of a numerical simulation is denoted $\Gamma = [c_i]_{0 \leq i \leq n-1}$, such that $\forall c_i,c_j \in \Gamma$, if $i \leq j$, then $c_i$ is computed before $c_j$.
\item Finally, a \textit{multi-stencil program} is defined by the septuplet $\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta, \mathcal{S},\Gamma)$.}
\end{itemize}
%!!!!!!!  est-ce qu'on fusionne le maillage, les entites et les domaines ?????
%\end{mydef}

>>>>

To numerically solve a set of PDEs, iterative methods (finite difference, finite volume or finite element methods) are frequently used to approximate the solution through a discretized (step by step) phenomena. Thus, the continuous time and space domains are discretized so that a set of numerical computations are iteratively (time discretization) applied onto a mesh (space discretization). In other words, the PDEs are transformed to a set of numerical computations applied at each time step on elements of the discretized space domain. Among those numerical computations are found a set of numerical schemes, also called \textit{stencil computations}, a set of auxiliary computations also called \emph{local computations}, and finally a set of \emph{reductions} to reduce a bunch of values to a single scalar value.
This section gives formal definitions of what we call a \textit{stencil program} and its computations. Then, the different mid-grain parallelization techniques which can be applied on such program, are presented.

@@

\begin{mydef}
A mesh is a connected undirected graph $\mathcal{M}=(V,E)$, where $V\subset \Omega$ is the set of vertices and $E\subseteq V^2$ the set of edges. The set of edges $E$ of a mesh $\mathcal{M}=(V,E)$ does not contain bridges. It is said that the mesh is applied onto $\Omega$.
\end{mydef}
\begin{figure}[!h]\begin{center}
  \resizebox{8cm}{!}{\includegraphics{./images/maillages.pdf}}
  \caption{From left to right, Cartesian, curvilinear and unstructured meshes.}
  \label{fig:mesh}
\end{center}\end{figure}
\begin{mydef}
The dimension of a mesh $\mathcal{M}=(V,E)$ applied onto $\Omega=\mathbb{R}^n$ is denoted $dim(\mathcal{M})=n$.
\end{mydef}
A mesh can be structured (as Cartesian or curvilinear meshes), unstructured, regular or irregular (without the same topology for each element) and hybrid as illustrated in Figure~\ref{fig:mesh}.

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item An entity $e$ of a mesh $\mathcal{M}=(V,E)$ is defined as a subset of its vertices and edges, $e\subset (V\cup E)$.
\item A mesh entities group is denoted $E$ such that $E \in \mathcal{P}(V\cup E)$. It represents a set of entities of the same type.
\item Finally the set of mesh entities groups of a simulation is denoted $\mathcal{E}$.
\end{itemize}
%\end{mydef}

@@

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item We denote a scalar as an identifier associated to a numerical value and applied onto a mesh $\mathcal{M}$ where $dim(\mathcal{M})=0$. In other words a scalar can be seen as a variable containing a numerical value. The set of scalars is denoted $\mathcal{S}$.
\item $\mathcal{T}=\mathbb{R}$ is the continuous time domain of a numerical simulation.
\item The discretization of the continuous time domain $\mathcal{T}$ is denoted as the pair $T(\Delta t,conv)$.
\begin{itemize}
\item $\Delta t \in \mathcal{S}$ represents the time interval of the numerical simulation, such that for a current time iteration $t_i$, the next time iteration is $t_{i+1} = t_i + \Delta t$. The default value of $\Delta t$ is $1$.
\item $conv:\mathcal{S}^n \rightarrow bool$ is a function which returns a boolean from a set of scalar variables. This function represents the convergence criteria of the simulation. 
\item At a given time step, the convergence criteria is evaluated such that if $conv$ returns $true$ the next time step can start such that $t_{i+1} = t_i + \Delta t$, otherwise the simulation ends. For example, the simplest $conv$ is typically to have two scalars: $t$ the time, and a fix number of iterations denoted $it$. At each time step $\Delta t=1$, and $conv$ returns the boolean expression $t<it$.
\end{itemize}
\end{itemize}
%\end{mydef}

@@

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item \textit{A quantity (or data) is a function $\delta: E_{\delta} \mapsto V_{\delta}$ which associates each entity of a group $e\in E_{\delta}$ to a value $v\in V_{\delta}$ where $V_{\delta}$ is a data type, typically $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.
\item The set of quantities applied onto the mesh is denoted $\Delta$.
\item In the rest of this paper, the group of mesh entities on which a quantity $\delta$ is mapped is denoted $entity(\delta)=E_{\delta}$.}
\end{itemize}
%\end{mydef}

@@

In this section are considered three different types of kernel computations, stencil kernels, local kernels, and reduction kernels.

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A computation domain $D$ is a subpart  a mesh entities group, $D \subset E \in \mathcal{E}$.
\item The set of computation domains of a numerical simulation is denoted $\mathcal{D}$.
\item A neighborhood $n$ is a function which for a given entity $e \in E_i$, returns a set of $m$ entities in $E_j$, $n : E_i \rightarrow E_j^m$. One can notice that $i = j$ is possible.
\item The set of neihborhood functions in a numerical simulation is denoted $\mathcal{N}$.
\item A single data is written during a computation and is denoted $w \in \Delta$. Thus, the computation domain of the computation is $D \subset E_w=entity(w)$.
\item A set of data to read and use in a computation is denoted $R$. Each element of $R$ is a tuple $(r,n)$, where $r \in \Delta$ and $n$ is a neighborhood function such that $n : E_w \rightarrow entity(r)^m$. The neighborhood indicates which entities, on which $r$ is applied, will be read during the computation.
\item A numerical expression $\text{exp}_{w}: entity(w) \times R^n \rightarrow V_{w}$ is a function which computes the values of the written quantity $w \in \Delta$ for a given entity $e \in entity(w)=E_w$, using a set of input data $R$.
\item A computation kernel $k$ of a numerical simulation is defined as $c(R,w,\text{exp},D)$, where $R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
\end{itemize}
%\end{mydef}
%If the number of computations in $\Gamma$ is $card(\Gamma)=n$, such that $\bigcup_{i=0}^{n-1}c_i = \Gamma$, then $\bigcup_{i=0}^{m-1}R_i \cup w_i \subseteq \Delta$.

@@

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item We denote by $identity$ the identity function $x=x$.
\item A kernel computation $k(R,w,\text{exp},D) \in \Gamma$ is a stencil kernel $\iff \exists (r,n) \in R$ such that $n \neq identity$.
\item A kernel computation $k(R,w,\text{exp},D) \in \Gamma$ is a local kernel $\iff \forall (r,n) \in R$, $n = identity$.
\end{itemize}

\noindent \textbf{Property}
A kernel for which all data read and written are applied onto a mesh of dimension $0$ is a local kernel.

@@

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A reduction computation $r$ of a numerical simulation is defined as $r(R,w,red)$, where $R$ is the set of data read, $w \in \mathcal{S}$ is the scalar written, and $\text{red}$ is an associative reduction function.
% \item For any $i,j$ such that $r_i \in R$ and $r_j \in R$, if $dim(mesh(r_i))=dim(mesh(r_j))=n$.
% \item The data written by the reduction $w$ verifies $dim(mesh(w))<n$.
\item A reduction function $\text{red}: \Delta^n \rightarrow \mathcal{S}$ is an operation which reduces a set of data in $\Delta$ (on it sentities) to a single scalar variable in $\mathcal{S}$. The reduction function must be associative.
\item The set of $n$ ordered computations (kernel or reduction) of a numerical simulation is denoted $\Gamma = [c_i]_{0 \leq i \leq n-1}$, such that $\forall c_i,c_j \in \Gamma$, if $i \leq j$, then $c_i$ is computed before $c_j$.
\item Finally, a \textit{multi-stencil program} is defined by the septuplet $\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta, \mathcal{S},\Gamma)$.
\end{itemize}
%!!!!!!!  est-ce qu'on fusionne le maillage, les entites et les domaines ?????
%\end{mydef}
